---
title: "STAT545_hw02_JLB"
output: github_document
---

#STAT 545 Homework 2

The different sections of this homework are broken down into the sections outlined in [the homework instructions](http://stat545.com/Classroom/assignments/hw02/hw02.html): 

**Homework 2 Task list: **
1. -[] Getting Started: Install Gapminder & dyplyr
2. -[]Smell test the data: Explore the gapminder object
3. -[]Explore individual variables:
4. -[]Explore various plot types: 
5. -[]Extra exercise: 
6. -[]Conclusions:


###Getting Started:
Install/Load Gapminder and dyplyr (through the tidyverse package)
```{r}
library(gapminder)
library(tidyverse)
```


###Smell test the data:
Explore the gapminder object:

**1. Is it a dataframe, matrix, vector, list?**
The typeof() function will tell me what data types gapminder contains. 
The class() function will tell me the class of the object gapminder.
```{r}
?gapminder
typeof(gapminder)
class(gapminder)
```
Conclusion: 
- typing ?gapminder in R returns information about the gapminder data. 
- Gapminder is a data frame that contains list data.

**2. What is its class?**

As shown above, we can use the class() function to determine the class of the gapminder object:
```{r}
?gapminder
(class(gapminder))
```
Conclusion: 
The class of gapminder is a data frame, specifically, a tibble (which we have not yet explored in class)

**3. How many columns/variables?**
```{r}
(ncol(gapminder))
```
Conclusion: 
There are 6 variables in gapminder.

**4. How many rows/observations?**
```{r}
(nrow(gapminder))
```
Conclusion: 
There are 1704 observations in gapminder.


**5. Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?**
I found additional information on how to further explore gapminder [here](http://adv-r.had.co.nz/Data-structures.html)
```{r}
?gapminder
length(gapminder) 
attributes(gapminder)

```
Conclusion: 

a. using ?gapminder is another quick way to get information about gapminder including: the format, the names and types of variables inside the object, and specific descriptions of each variable. 
context: This function would be useful for getting a quick overview of the object, as well as in the context of needing help with understanding the data and where it comes from.

b. Using length(gapminder) will return how many elements are inside it, in this case, returns the number of variables. Another application of the length() function would become useful when looking at the length of variables such as an array. This function would also be useful for creating a loop that repeats for each value within a variable.

c. the attributes(gapminder) function is similar to the summary(), but it gives information about the row names as well. In this context the row names are simply numbers 1, 2, 3, etc..., but in other contexts these names may be more useful. 


**What data type is each variable?**
using the str() function allows us to look at the structure of the variables 

```{r}

str(gapminder)
#to double check the output of str(), I can check a variable individually:
class(gapminder$year)
  
```
Conclusion: 
The variables in gapminder are lists containing different datatypes:
country - factors
continent - factors
year - integers
lifeExp - numbers
pop - integers
gdpPercap - numbers

####Explore individual variables

I am curious about a few questions regarding this data, notably, the following questions:

1. What is the relationship between life expectancy and GDP per capita?
2. How did life expectancy in the continent of Africa change over time?

In order to address these questions, I will need to consider several variables:

1. Is there a relationship between lifeExpectancy and gdpPercap?
    Life expectancy, GDP per capita. I will also need to consider a particular year. 
    
2. How did life expectancy in the continent of Africa change over time?
    To answer this question, I will need to look at the continent variable as well as the year variable.

What are possible values (or range, whichever is appropriate) of each variable?
What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand.
Feel free to use summary stats, tables, figures. We’re NOT expecting high production value (yet).

```{r}
min(gapminder$year)
max(gapminder$year)
range(gapminder$year)
summary(gapminder$continent)
```

####Explore various plot types
 - A scatterplot of two quantitative variables.
 - A plot of one quantitative variable. Maybe a histogram or densityplot or frequency polygon.
 - A plot of one quantitative variable and one categorical. Maybe boxplots for several continents or countries.
```{r}

```


####Extra Exercise:

Instructions: Evaluate this code and describe the result. Presumably the analyst’s intent was to get the data for Rwanda and Afghanistan. Did they succeed? Why or why not? If not, what is the correct way to do this?
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

