---
title: "STAT545_hw02_JLB"
output: github_document
---

#STAT 545 Homework 2

The different sections of this homework are broken down into the sections outlined in [the homework instructions](http://stat545.com/Classroom/assignments/hw02/hw02.html): 

**Homework 2 Task list: **
1. -[] Getting Started: Install Gapminder & dyplyr
2. -[]Smell test the data: Explore the gapminder object
3. -[]Explore individual variables:
4. -[]Explore various plot types: 
5. -[]Extra exercise: 
6. -[]Conclusions:


###Getting Started:
Install/Load Gapminder and dyplyr (through the tidyverse package)
```{r}
library(gapminder)
library(tidyverse)
```


###Smell test the data:
Explore the gapminder object:

**1. Is it a dataframe, matrix, vector, list?**
The typeof() function will tell me what data types gapminder contains. 
The class() function will tell me the class of the object gapminder.
```{r}
?gapminder
typeof(gapminder)
class(gapminder)
```
Conclusion: 
- typing ?gapminder in R returns information about the gapminder data. 
- Gapminder is a data frame that contains list data.

**2. What is its class?**

As shown above, we can use the class() function to determine the class of the gapminder object:
```{r}
?gapminder
(class(gapminder))
```
Conclusion: 
The class of gapminder is a data frame, specifically, a tibble (which we have not yet explored in class)

**3. How many columns/variables?**
```{r}
(ncol(gapminder))
```
Conclusion: 
There are 6 variables in gapminder.

**4. How many rows/observations?**
```{r}
(nrow(gapminder))
```
Conclusion: 
There are 1704 observations in gapminder.


**5. Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?**
I found additional information on how to further explore gapminder [here](http://adv-r.had.co.nz/Data-structures.html)
```{r}
?gapminder
length(gapminder) 
attributes(gapminder)

```
Conclusion: 

a. using ?gapminder is another quick way to get information about gapminder including: the format, the names and types of variables inside the object, and specific descriptions of each variable. 
context: This function would be useful for getting a quick overview of the object, as well as in the context of needing help with understanding the data and where it comes from.

b. Using length(gapminder) will return how many elements are inside it, in this case, returns the number of variables. Another application of the length() function would become useful when looking at the length of variables such as an array. This function would also be useful for creating a loop that repeats for each value within a variable.

c. the attributes(gapminder) function is similar to the summary(), but it gives information about the row names as well. In this context the row names are simply numbers 1, 2, 3, etc..., but in other contexts these names may be more useful. 


**What data type is each variable?**
using the str() function allows us to look at the structure of the variables 

```{r}

str(gapminder)
#to double check the output of str(), I can check a variable individually:
class(gapminder$year)
  
```
Conclusion: 
The variables in gapminder are lists containing different datatypes:
country - factors
continent - factors
year - integers
lifeExp - numbers
pop - integers
gdpPercap - numbers

####Explore individual variables

####Exploring the year variable:
I can look at the range of years contained in the data:
The min() function shows the earliest year, while the max() function shows the latest year. The range() function gives both min and max as an output (ie, the range of the years). 

```{r}
min(gapminder$year)
max(gapminder$year)
range(gapminder$year)
```
**Conclusion:**
the range of years in the gapminder dataset is 1952-2007.

Looking at the spread of values :
```{r}
#spread(gapminder, key,value)
```

I can also use the min(), max() and range() functions with the lifeExp, because it contains integers, however, this will not work for the country variable, as these functions do not apply.
```{r}
min(gapminder$lifeExp)
max(gapminder$lifeExp)
range(gapminder$lifeExp)

```
Conclusion: 
The range of life expectancies contained in the gapminder dataset is 23.6 - 82.6 years.

To look at the distribution of the continent variable:
```{r}
attributes(gapminder$continent)
```
These values seem typical, as they are representative of the world's continents.




SUMMARY TABLE:



Looking more closely at the country and lifeExp variables: 
```{r}
select(gapminder, country, year, lifeExp) %>% 
  filter(year == "2007") %>% 
  summary()

```

```{r}

```


####Explore various plot types
 useful [site](https://stackoverflow.com/questions/10438752/adding-x-and-y-axis-labels-in-ggplot2) for how to add axis labels and title for your ggplot

 - A scatterplot of two quantitative variables.

```{r}
#similar to how we saw filtering by a variable such as 'country' in class for geom_line, we can also do this to assign colours using colour = country. 

gapminder %>% 
  
  filter(country == "Canada"|country == "United States") %>% 
  ggplot(aes(x=year, y=pop)) + 
  geom_point() +
  geom_line(aes(filter = country,colour = country)) +
  xlab("Year") +
  ylab("Population of Canada and United States") +
  ggtitle("North American Population Trends")

  

```


```{r}
gapminder %>% 
  filter(country == "Canada" | country =="United States")%>% 
  ggplot(aes(x=year, y = gdpPercap)) +
  geom_line(aes(group = country, colour = country)) +
  geom_point() +
  xlab("Year") +
  ylab("GDP Per Capita") +
  ggtitle("North American GDP Per Capita Trends")
  
```



- A plot of one quantitative variable. Maybe a histogram or densityplot or frequency polygon.

```{r}
gapminder %>%
  filter (year == "2007") %>% 
  ggplot(aes(gdpPercap)) +
  geom_histogram(aes(y=..density..),bins = 35) + 
  geom_density(fill = "green", alpha = 0.2)


#looking more closely at the spread of density where gdpPercap is less than 10 000:

gapminder 
```



```{r}
gapminder %>%
  select(gdpPercap, continent) %>% 
  ggplot(aes(continent, gdpPercap)) +
  scale_y_log10() +
  geom_violin() +
  geom_jitter(alpha = 0.15) +
  xlab("Continent")+
  ylab("GDP Per Capita (log10)") +
  ggtitle("GDP Per Capita by Continent")
  
```

 - A plot of one quantitative variable and one categorical. Maybe boxplots for several continents or countries.
####Extra Exercise:

Instructions: Evaluate this code and describe the result. Presumably the analyst’s intent was to get the data for Rwanda and Afghanistan. Did they succeed? Why or why not? If not, what is the correct way to do this?
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

