---
title: "STAT545_hw02_JLB"
output: github_document
---

#STAT 545 Homework 2

The different sections of this homework are broken down into the sections outlined in [the homework instructions](http://stat545.com/Classroom/assignments/hw02/hw02.html): 

**Homework 2 Task list: **  

1. Getting Started: Install Gapminder & dyplyr    
2. Smell test the data: Explore the gapminder object    
3. Explore individual variables    
4. Explore various plot types     
5. Extra exercise   
6. Conclusions and Reflection    


### 1. Getting Started:   
Install/Load Gapminder and dyplyr (through the tidyverse package)
```{r Load Gapminder and dyplyr, message = FALSE}
library(gapminder)
library(tidyverse)
```


### 2. Smell test the data:
Explore the gapminder object:

**2a. Is it a dataframe, matrix, vector, list?**  
The typeof() function will tell me what data types gapminder contains.   
The class() function will tell me the class of the object gapminder.
```{r 2a: What is Gapminder?} 
typeof(gapminder)
class(gapminder)
```
Conclusion:  
- Gapminder is a data frame that contains list data.

**2b. What is its class?**  
As shown above, we can use the class() function to determine the class of the gapminder object:
```{r Class gapminder}
(class(gapminder))
```
Conclusion:   
- The class of gapminder is a data frame, specifically, a tibble (which we have not yet explored in class)

**2c. How many columns/variables?**
```{r columns gapminder}
(ncol(gapminder))
```
Conclusion:   
- There are 6 variables in gapminder.

**2d. How many rows/observations?**
```{r rows gapminder}
(nrow(gapminder))
```
Conclusion:   
- There are 1704 observations in gapminder.

**2e Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?**  
I found additional information on how to further explore gapminder [here](http://adv-r.had.co.nz/Data-structures.html)
```{r Further gapminder exploration}
dim(gapminder) #returns the number of rows and columns, respectively.
length(gapminder) #returns the number of columns
levels(gapminder$continent) #returns the groups in a particular variable
```
Conclusion:  
- using ?gapminder in the console is another way to get information about gapminder such as format, names and types of variables in the object, and specific descriptions of each variable.   
context: This function is useful for getting a quick overview and understanding the data and where it comes from.  
- length(gapminder) returns the number of elements inside it. The length() function might be used to make a loop that repeats for each value in the array.  
- the levels(), attributes(), and summary() functions can provide useful information about the variables inside a dataset, as well as basic statistical information about these.


**2f What data type is each variable?**  
using the str() function allows us to look at the structure of the variables 
```{r data types in gapminder}
str(gapminder)
#to double check the output of str(), I can check a variable individually:
class(gapminder$year)
```
Conclusion:  
The variables in gapminder are lists containing different datatypes:  
- country - factors  
- continent - factors  
- year - integers  
- lifeExp - numbers  
- pop - integers  
- gdpPercap - numbers  

#### 3. Explore individual variables
Categorical variable selected: continent  
Quantitative variable selected: gdpPercap

**3a. Exploring possible values (or range, whichever is appropriate) of variables:**    

```{r range of values of continent and gdpPercap}
min(gapminder$gdpPercap) #returns minimum value of gdpPercap
max(gapminder$gdpPercap) #returns maximum value of gdpPercap
range(gapminder$gdpPercap) #returns the range lowest and highest values of gdpPercap
levels(gapminder$continent)#returns the groups or levels contained in the continent variable

```
Conclusion:   
- The range of GDP per capita in the gapminder dataset is $241-113 523.  
- The possible values of the continent variable are: Africa, Americas, Asia, Europe, and Oceania.
- By inspection, the range of values contained in GDP per capita and the possible values contained in the continent variable seem appropriate.


#### 4. Explore various plot types

 **4a. A scatterplot of two quantitative variables.**  
 
I will compare population growth in Canada and the United States from 1952-2007.

```{r Explore Plot Types - Scatterplot}
gapminder_CAN_USA = gapminder %>% 
  filter(country == "Canada"|country == "United States")
  
  ggplot(gapminder_CAN_USA, aes(x=year, y=pop)) + 
  geom_point() +
  geom_line(aes(colour = country)) +
  xlab("Year") +
  ylab("Population of Canada and United States") +
  ggtitle("North American Population Trends")

```

I will compare GDP per Capita in Canada and the United States from 1952-2007.

```{r Explore gdpPercap trends}

  ggplot(gapminder_CAN_USA,aes(x=year, y = gdpPercap)) +
  geom_line(aes(group = country, colour = country)) +
  geom_point() +
  xlab("Year") +
  ylab("GDP Per Capita") +
  ggtitle("North American GDP Per Capita Trends")
  
```

  
**4b. A plot of one quantitative variable.**  

 I will make a histogram of GDP per capita in 2007. 
 I will first assign the filtered gapminder dataframe for the year 2007, as I will use this for subsequent plots.
```{r Histogram Plot quantitative }
gapminder_2007 = gapminder %>%
  filter (year == "2007")

  ggplot(gapminder_2007, aes(gdpPercap)) +
  geom_histogram(aes(y=..density..),bins = 35) + 
  geom_density(fill = "red", alpha = 0.2) +
  scale_x_log10() +
  xlab("GDP Per Capita") +
  ylab("Density") +
  ggtitle("Histogram of GDP Per Capita in 2007")
```
  
    
**4c. A plot of one quantitative variable and one categorical.**

I will plot the gdp per capita of various continents in 2007.
```{r GDP per capita by continent}

  ggplot(gapminder_2007, aes(continent, gdpPercap)) +
  scale_y_log10() +
  geom_violin(fill = "darkslategray1", alpha = 0.5) +
  geom_jitter(alpha = 0.33) +
  xlab("Continent")+
  ylab("GDP Per Capita") +
  ggtitle("GDP Per Capita by Continent in 2007")
  
```


**4d. Exploring other dyplr functions beyond filter() and select()**

**The mutate() function:**    

The mutate function can be used to create a new variable by making changes to existing data.  
Here I will multiply the GDP per capita by the population to examine the GDP of countries, organized by continent.
```{r}
gapminder_2007 %>% 
  mutate(GDP = gdpPercap * pop) %>% 
  ggplot(aes(continent, GDP)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.2) +
  scale_y_log10() +
  ylab("Gross Domestic Product") +
  xlab("Continent") +
  ggtitle("GDP per Capita in 2007 by Continent")
```
 
**The between() function: **  

The between() function selects for values of a variable between two assigned values.   
Here I will look at the countries with a GDP Per capita between two values for the year 2007. 
```{r}
gapminder_2007 %>% 
  filter (between(gdpPercap, 30000,50000) ) %>% 
  ggplot(aes(gdpPercap)) + geom_histogram(aes(fill = continent), bins = 10) +
  facet_wrap(~continent)+
  ylab("Count") +
  xlab("GDP per capita") +
  ggtitle("Countries with GDP per capita between 30 and 50k")

```

**The top_n() function **   

The top_n() function can be used to select the top (or bottom) values in a data set. 
Here I will show two tibbles showing the countries with the lowest and highest GDP per capita in 2007.
```{r}
#Table showing the 5 countries with the lowest GDP per capita in 2007, in ascending order. 
gapminder_2007 %>% 
  select (country, gdpPercap) %>% 
  top_n (-5) %>% 
  arrange(gdpPercap)

#Table showing the 5 countries with highest GDP per capita in 2007 in ascending order
gapminder_2007 %>% 
  select (country, gdpPercap) %>% 
  top_n (5) %>% 
  arrange(gdpPercap)
```


####5. Extra Exercise:

Instructions: Evaluate this code and describe the result. Presumably the analyst’s intent was to get the data for Rwanda and Afghanistan. Did they succeed? Why or why not? If not, what is the correct way to do this?

Result: No the analyst was not successful. 
```{r Extra Exercise}
filter(gapminder, country == c("Rwanda", "Afghanistan"))

#looking at the output, the analyst's code returns 12 rows of data (6 for Rwanda, 6 for Afghanistan).   
#However, when I run the code below I get 24 rows of data.

gapminder %>% 
  filter (country =="Rwanda" | country == "Afghanistan")


```
When the analyst used c(), they ommitted half of the rows.   
Why?  
When we use c(), the code it will go row by row through the different countries it will compare row1 with "Rwanda".  
Next, it will compare row2 with "Afghanistan".  
It will then start over and evaluate row 3 == "Rwanda", and row 4 =="Afghanistan", etc.   
If the result is TRUE, it will add it to the tibble.   
Therefore, when comparing to "Rwanda" it will return FALSE if the row contains "Afghanistan" and vice versa  


To test if the above explanation is correct:  

If I run the code below, I should obtain 8 rows with values for Rwanda, but only 4 for Afghanistan   
**Note** - this is only applicable for this dataset, as I know there are equal numbers of rows for Rwanda and Afghanistan, and I know they are organized alphabetically. 
```{r Extra Exercise test}
filter(gapminder, country == c("Rwanda", "Afghanistan", "Rwanda"))
```

Interestingly, If I run the code shown below, it returns a warning.  

```{r wrong filtering}
filter(gapminder, country == c("Rwanda", "Afghanistan", "Norway", "China", "Japan")) 

```
The reason for the warning is because the list of countries is not fully cycled through, as we cannot divide 5 into the number of rows of data in gapminder. 

Conclusion:    
- A correct way to filter the data without accidentally leaving out rows of data is shown below:
```{r Correct filtering, results="hide" }
gapminder %>% 
  filter (country =="Rwanda" | country == "Afghanistan")
```








