---
title: "STAT545_hw02_JLB"
output: github_document
---
The different sections of this homework are broken down into the sections outlined in [the homework instructions](http://stat545.com/Classroom/assignments/hw02/hw02.html): 

Task list: 
1. -[] Getting Started: Install Gapminder & dyplyr
2. -[]Smell test the data: Explore the gapminder object
3. -[]Explore individual variables:
4. -[]Explore various plot types: 
5. -[]Extra exercise: 
6. -[]Conclusions:



####Getting Started:
Install/Load Gapminder and dyplyr (through the tidyverse package)
```{r}
library(gapminder)
library(tidyverse)
```


####Smell test the data:
Explore the gapminder object:

**1. Is it a dataframe, matrix, vector, list?**
typeof() will tell me what data types gapminder contains. 
```{r}
?gapminder
?typeof()
typeof(gapminder)
class(gapminder)
```
Conclusion: typing ?gapminder in R returns information about the gapminder data. It is a data frame that contains list data.

**2. What is its class?**

We can use the class() function to determine the class of the gapminder object:
```{r}
?gapminder
(class(gapminder))
```
Conclusion: The class of gapminder is a data frame, specifically, a tibble (which we have not yet explored in class)

**3. How many columns/variables?**
```{r}
(ncol(gapminder))
```
Conclusion: There are 6 different variables in gapminder.

**4. How many rows/observations?**
```{r}
(nrow(gapminder))
```
Conclusion: There are 1704 observations in gapminder.


**5. Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?**
I found additional information on how to further explore gapminder [here](http://adv-r.had.co.nz/Data-structures.html)
```{r}
?gapminder
length(gapminder) #using length(gapminder) will return how many elements inside it (in this case, it contains 6 elements aka variables, each in the form of a list)
attributes(gapminder)


```
Note: using ?gapminder is another quick way to get information about gapminder including: the format, the names and types of variables inside the object, and specific descriptions of each variable. 
This function would be useful for getting a quick overview of the object, as well as useful in the context of needing help with understanding the data and where it comes from.



**What data type is each variable?**
```{r}
str(gapminder)
```

```{r}
(typeof(gapminder))
(str(gapminder))
(ncol(gapminder))
(nrow(gapminder))
(summary(gapminder))
(class(gapminder))
```

####Explore individual variables
```{r}

```

####Explore various plot types
 - A scatterplot of two quantitative variables.
 - A plot of one quantitative variable. Maybe a histogram or densityplot or frequency polygon.
 - A plot of one quantitative variable and one categorical. Maybe boxplots for several continents or countries.
```{r}

```


####Extra Exercise:

Instructions: Evaluate this code and describe the result. Presumably the analyst’s intent was to get the data for Rwanda and Afghanistan. Did they succeed? Why or why not? If not, what is the correct way to do this?
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

